---
    title: "Lab: Multinomial Regression"
    subtitle: "Princeton University"
    date: "2025-02-26"
    author: "KW"
    categories: [code, analysis]
    output: 
      tufte::tufte_html:
        css: 
        tufte_variant: "envisioned"
        highlight: github-dark
        fig_height: 10
        fig_width: 16
        toc: true
        toc_depth: 1
    format:
      html:
        self-contained: false
        anchor-sections: true
        code-tools: true
        code-fold: true
        fig-width: 8
        fig-height: 4
        code-block-bg: "#f1f3f5"
        code-block-border-left: "#31BAE9"
        mainfont: Source Sans Pro
        theme: journal
        toc: true
        toc-depth: 3
        toc-location: left
        captions: true
        cap-location: margin
        table-captions: true
        tbl-cap-location: margin
        reference-location: margin

    comments:
      hypothesis: false

    execute:
      warning: false
      message: false
    engine: knitr
---

Lab Goal: Predict voting frequency using demographic variables Data source: FiveThirtyEight "Why Many Americans Don't Vote" survey Method: Multinomial logistic regression

## Data

The data for this assignment comes from an online Ipsos survey that was conducted for the FiveThirtyEight article ["Why Many Americans Don't Vote"](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/). You can read more about the survey design and respondents in the README of the [GitHub repo](https://github.com/fivethirtyeight/data/tree/master/non-voters) for the data.

Respondents were asked a variety of questions about their political beliefs, thoughts on multiple issues, and voting behavior. We will focus on using the demographic variables and someone's party identification to understand whether a person is a probable voter.

The variables we'll focus on were (definitions from the codebook in data set GitHub repo):

-   `ppage`: Age of respondent

-   `educ`: Highest educational attainment category.\

-   `race`: Race of respondent, census categories. Note: all categories except Hispanic were non-Hispanic.

-   `gender`: Gender of respondent

-   `income_cat`: Household income category of respondent

-   `Q30`: Response to the question "Generally speaking, do you think of yourself as a..."

    -   1: Republican
    -   2: Democrat
    -   3: Independent
    -   4: Another party, please specify
    -   5: No preference
    -   -1: No response

-   `voter_category`: past voting behavior:

    -   **always**: respondent voted in all or all-but-one of the elections they were eligible in
    -   **sporadic**: respondent voted in at least two, but fewer than all-but-one of the elections they were eligible in
    -   **rarely/never**: respondent voted in 0 or 1 of the elections they were eligible in

You can read in the data directly from the GitHub repo:

```{r, message=FALSE}
library(nnet)
library(car)
library(tidyverse)
library(emmeans)
library(ggeffects)
library(knitr)
library(patchwork)
library(broom)
library(parameters)
library(easystats)
```

```{r}

voter_data <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv")
```

# Lab

-   The variable `Q30` contains the respondent's political party identification. Make a new variable that simplifies `Q30` into four categories: "Democrat", "Republican", "Independent", "Other" ("Other" also includes respondents who did not answer the question).

```{r}
voter_data <- voter_data %>%
  mutate(pol_ident_new = case_when(
    Q30==1 ~ "Rep", 
    Q30==2 ~ "Dem", 
    Q30==3 ~ "Indep", 
    TRUE ~ "Other"
  ))
```

-   The variable `voter_category` identifies the respondent's past voter behavior. Relevel the variable to make rarely/never the baseline level, followed by sporadic, then always

```{r}
voter_data <- voter_data %>%
  mutate(voter_category = factor(voter_category,
                           levels = c("rarely/never", "sporadic", "always"),
                           ordered = TRUE))

```

-   Center the age variable to make the intercept more interepretable. That is, so that it reflects the log-odds for an average-aged person rather than a 0-year old person

```{r}
voter_data = voter_data %>%
  mutate(ppage = scale(voter_data$ppage,scale=FALSE))
```

-   In the [FiveThirtyEight article](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/), the authors include visualizations of the relationship between the voter category and demographic variables such as race, age, education, etc. Select two demographic variables. For each variable, try to replicate the visualizations and interpret the plot to describe its relationship with voter category. Have fun with it: https://www.mikelee.co/posts/2020-02-08-recreate-fivethirtyeight-chicklet-stacked-bar-chart-in-ggplot2.

```{r}
# library
library(ggplot2)
library(viridis)
library(cowplot)
library(ggchicklet)

# Enter code
voter_race <- voter_data %>%
  count(race, voter_category) %>%
  mutate(proportion = n / sum(n))

# ggplot(voter_race, aes(x = race, y = proportion, fill = voter_category)) +
#  geom_bar(stat="identity",position="fill") + coord_flip()
#   labs(x="Race", y="Proportion")

plt_race = voter_race %>%
  ggplot(aes(x=race, y=proportion, group = voter_category, fill = voter_category)) +
  geom_chicklet(width = 0.75, position="fill")  + 
  scale_y_continuous(
    expand = c(0, 0.0625),
    position = "right",
    breaks = seq(0, 1, 0.25),
    labels = c("0%", seq(25, 100, 25))
  ) +
  coord_flip() + 
  labs(x="Race", y="Proportion")
plt_race
```

```{r}
# Enter code
voter_gender <- voter_data %>%
  count(gender, voter_category) %>%
  mutate(proportion = n / sum(n))

plt_gender = voter_gender %>%
  ggplot(aes(x=gender, y=proportion, group = voter_category, fill = voter_category)) +
  geom_chicklet(width = 0.75, position="fill")  + 
  scale_y_continuous(
    expand = c(0, 0.0625),
    position = "right",
    breaks = seq(0, 1, 0.25),
    labels = c("0%", seq(25, 100, 25))
  ) +
  coord_flip() + 
  labs(x="Gender", y="Proportion")
plt_gender
```

The plots can be combined into a single plot using the patchwork package.

```{r}
library(patchwork)
plt_race / plt_gender
```

-   Fit a model using mean-centered age, race, gender, income, and education to predict voter category. Show the code used to fit the model, but do **not** display the model output.

```{r}
library(nnet)

model = multinom(voter_category ~ ppage + race + gender + income_cat + educ, 
                     data = voter_data)

```

-   *Should party identification be added to the model?*
-   #Hint: Use an anova test to make the determination

```{r}
model2 = multinom(voter_category ~ ppage + race + gender + income_cat + educ + pol_ident_new, 
                     data = voter_data)

anova(model, model2) %>% 
  kable()
```

```         
> #Enter answer based on your code: Yes, party identification should be included in the model
```

**Use the model you select for the remainder of the assignment**.

## LRT

-   Run the full model and report overall significance of each of the terms

    ```{r}
    car::Anova(model2) %>% 
      kable()
    ```

    > All the terms: age, race, gender, income, education, and party identification are all significant in predicting voter category

## Marginal Effects Political Group - Emmeans

```{r}
#Get estimated marginal means from the model

#using 
multinomial_analysis <- emmeans(model2, ~ pol_ident_new|voter_category)

coefs = contrast(regrid(multinomial_analysis, "log"),"trt.vs.ctrl1",  by="pol_ident_new")
# you can add a parameter to the above command, ref = newbaseline, if you want to change baseline

update(coefs, by = "contrast") %>%
 kable(format = "markdown", digits = 3)
```

## Marginal Effects of Education - Emmeans

```{r}
#Get estimated marginal means from the model

#using 
multinomial_analysis <- emmeans(model2, ~ educ|voter_category)

coefs = contrast(regrid(multinomial_analysis, "log"),"trt.vs.ctrl1",  by="educ")
# you can add a parameter to the above command, ref = newbaseline, if you want to change baseline

update(coefs, by = "contrast") %>%
 kable(format = "markdown", digits = 3)
```

-   Next, plot the predicted probabilities of voter category as a function of Age and Party ID

```{r}
  ggemmeans(model2, terms = c("ppage")) %>% 
      ggplot(., aes(x = x, y = predicted, fill = response.level)) +
      geom_area() + 
      geom_rug(sides = "b", position = "jitter", alpha = .5) + 
      labs(x = "\nAge", y = "Predicted Probablity\n", title = "Predicted Probabilities of Voting Frequency by Age") +
      scale_fill_manual(
        name = NULL,
        values = c("always" = "#F6B533", "sporadic" = "#D07EA2", "rarely/never" = "#9854F7"),
        labels = c("RARELY OR NEVER VOTE    ", "SOMETIMES VOTE    ", "ALMOST ALWAYS VOTE    "),
        breaks = c("rarely/never", "sporadic", "always")
      ) +
      theme_minimal()
    
    
 ggemmeans(model2, terms=c("pol_ident_new")) %>%   ggplot(., aes(x = x, y = predicted, fill = response.level)) + 
  geom_bar(stat = "identity" ) +
    geom_text(aes(label = round(predicted, 3)), color="white", position = position_fill(vjust = 0.5),size=3)  + 
  labs(x="Education", y="Predicted Probablity") + 
  theme(text = element_text(size = 30)) +  
  scale_fill_viridis(discrete = TRUE) + 
  theme_lucid(base_size=25)
```

Plot predicted probabilities as a function of education and voting frequency.

```{r}
 ggemmeans(model2, terms=c("educ")) %>% ggplot(., aes(x = x, y = predicted, fill = response.level)) + 
  geom_bar(stat = "identity" ) +
    geom_text(aes(label = round(predicted, 3)), color="white", position = position_fill(vjust = 0.5),size=3)  + 
  labs(x="Education", y="Predicted Probablity") + 
  theme(text = element_text(size = 30)) +  
  scale_fill_viridis(discrete = TRUE) + 
  theme_lucid(base_size=25)
```

## Write-up

### Differences between political groups and voting behavior - Emmeans

```{r}
multinomial_analysis <- emmeans(model2, ~ pol_ident_new|voter_category)

coefs = contrast(regrid(multinomial_analysis, "log"),"trt.vs.ctrl1",  by="pol_ident_new")
# you can add a parameter to the above command, ref = newbaseline, if you want to change baseline

update(coefs, by = "contrast") %>%
 kable(format = "markdown", digits = 3)

# get difference between yes-no and fair-excellent
contrast(coefs, "revpairwise", by = "contrast") %>%
  kable(format = "markdown", digits = 3)
```

Enter your interpretation here:

Voters who are Democrats are 2.61 times more likely to vote sporadically than vote rarely/never.

Voters who are Democrats are 1.62 times more likely to vote always than vote rarely/never.

Voters who are Independents are 1.81 times more likely to vote sporadically than vote rarely/never.

Voters who are Independents are 4.78% less likely to vote always than vote rarely/never.

Voters who belong to other political parties are 1.08 times more likely to vote sporadically than vote rarely/never.

Voters who belong to other political parties are 56.61% less likely to vote always than vote rarely/never.

Voters who are Republicans are 2.42 times more likely to vote sporadically than vote rarely/never.

Voters who are Republicans are 1.39 times more likely to vote always than vote rarely/never.

Voters who are Independents are 30.93% less likely to vote sporadically than vote rarely/never compared to voters who are Democrats.

Voters who belong to voters who belong to other political parties are 58.65% less likely to vote sporadically than vote rarely/never compared to voters who are Democrats.

Voters who belong to other political parties are 40.13% less likely to vote sporadically than vote rarely/never compared to voters who are Republicans.

Voters who are Republicans are 7.50% less likely to vote sporadically than vote rarely/never compared to voters who are Democrats.

Voters who are Republicans are 1.34 times more likely to vote sporadically than vote rarely/never compared to voters who are Independents.

Voters who are Republicans are 2.24 times more likely to vote sporadically than vote rarely/never compared to voters who belong to other political parties.

Voters who are Independents are 41.08% less likely to vote always than vote rarely/never compared to voters who are Democrats.

Voters who belong to other political parties are 73.15% less likely to vote always than vote rarely/never compared to voters who are Democrats.

Voters who belong to other political parties are 54.43% less likely to vote always than vote rarely/never compared to voters who are Republicans.

Voters who are Republicans are 14.19% less likely to vote always than vote rarely/never compared to voters who are Democrats.

Voters who are Republicans are 1.46 times more likely to vote always than vote rarely/never compared to voters who are Independents.

Voters who are Republicans are 3.20 times more likely to vote always than vote rarely/never compared to voters who belong to other political parties.

### Differences between education level and voting behavior - Emmeans

Last part of the assignment: Interpret the results from running the following code for your model

```{r}
multi_an <- emmeans(model, ~ educ|voter_category)

coefs = contrast(regrid(multi_an, "log"),"trt.vs.ctrl1",  by="educ")

update(coefs, by = "contrast") %>% 
  kable(format = "markdown", digits = 3)

# get difference between yes-no and fair-excellent
contrast(coefs, "revpairwise", by = "contrast") %>%
  kable(format = "markdown", digits = 3)
```

Enter your interpretation here:

Voters with a highest degree of college are 3.18 times more likely to vote sporadically than vote rarely/never.

Voters with a highest degree of college are 2.04 times more likely to vote always than vote rarely/never.

Voters with a highest degree of high school or less are 1.30 times more likely to vote sporadically than vote rarely/never.

Voters with a highest degree of high school or less are 45.23% less likely to vote always than vote rarely/never.

Voters with a highest degree of some college are 2.24 times more likely to vote sporadically than vote rarely/never.

Voters with a highest degree of some college are 1.36 times more likely to vote always than vote rarely/never.

Voters with a highest degree of high school or less are 59.22% less likely to vote sporadically than vote rarely/never compared to voters with a highest degree of college.

Voters with a highest degree of some college are 29.46% less likely to vote sporadically than vote rarely/never compared to voters with a highest degree of college.

Voters with a highest degree of some college are 1.73 times more likely to vote sporadically than vote rarely/never compared to voters with a highest degree of high school or less.

Voters with a highest degree of high school or less are 73.10% less likely to vote always than vote rarely/never compared to voters with a highest degree of college.

Voters with a highest degree of some college are 33.03% less likely to vote always than vote rarely/never compared to voters with a highest degree of college.

Voters with a highest degree of some college are 2.49 times more likely to vote always than vote rarely/never compared to voters with a highest degree of high school or less.
